**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-05-15 submission/manuscript.Rmd
add changes par lab meeting
modulate and change

```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
######################################################################
# Load in needed functions and libraries
melt_data <-  function(data) {
  data_melt <- data %>%
    melt(measure.vars=c('cv_aucs', 'test_aucs')) %>%
    rename(AUC=value) %>%
    mutate(Performance = case_when(variable == "cv_aucs" ~ 'cross-validation', variable == "test_aucs" ~ 'testing')) %>%
    group_by(Performance)
  return(data_melt)
}

# Read in files as delim that are saved in a list with a pattern
read_files <- function(filenames){
  for(file in filenames){
    # Read the files generated in main.R
    # These files have cvAUCs and testAUCs for 100 data-splits
    data <- read.delim(file, header=T, sep=',')
  }
  return(data)
}

# Get model name with sub from file name
get_model_name <- function(files){
  pat1 <- "../data/process/walltime_"
  name_files <- sub(pat1, "", files)
  pat2 <- ".csv"
  names <- sub(pat2, "", name_files)
  return(names)
}
######################################################################

######################################################################
#----------------- Load OTU table and MetaData -----------------#
######################################################################

# Read in metadata and select only sample Id and diagnosis columns
meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
  select(sample, dx, Dx_Bin, fit_result)

# Read in OTU table and remove label and numOtus columns
shared <- read.delim('../data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
   select(-label, -numOtus)

# Merge metadata and OTU table.
# Group advanced adenomas and cancers together as cancer and normal, high risk normal and non-advanced adenomas as normal
# Then remove the sample ID column
data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
  mutate(dx = case_when(
    Dx_Bin== "Adenoma" ~ "normal",
    Dx_Bin== "Normal" ~ "normal",
    Dx_Bin== "High Risk Normal" ~ "normal",
    Dx_Bin== "adv Adenoma" ~ "cancer",
    Dx_Bin== "Cancer" ~ "cancer"
  )) %>%
  select(-sample, -Dx_Bin) %>%
  drop_na()
# We want the diagnosis column to be a factor
data$dx <- factor(data$dx)

######################################################################
#--------- Load .csv datato get mean test AUC for 7 models--------#
######################################################################

# Read in the cvAUCs, testAUCs for 100 splits.
best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
logit <- read_files(best_files[4])
l2svm <- read_files(best_files[3])
l1svm <- read_files(best_files[2])
rbf <- read_files(best_files[6])
rf <- read_files(best_files[5])
dt <- read_files(best_files[1])
xgboost <- read_files(best_files[7])

@@

rf_xgboost <- wilcox.test(rf$test_aucs, xgboost$test_aucs)
rf_l1svm <- wilcox.test(rf$test_aucs, l1svm$test_aucs)
l1svm_dt <- wilcox.test(l1svm$test_aucs, dt$test_aucs)
l1svm_rbf <- wilcox.test(l1svm$test_aucs, rbf$test_aucs)
l1svm_l2svm <- wilcox.test(l1svm$test_aucs, l2svm$test_aucs)
l1svm_logit <- wilcox.test(l1svm$test_aucs, logit$test_aucs)

# Get the unpaired two-samples Wilcoxon test to see if cv and test differ significantly
logit_diff <- wilcox.test(logit$test_aucs, logit$cv_aucs)
l2svm_diff <- wilcox.test(l2svm$test_aucs, l2svm$cv_aucs)
rbf_diff <- wilcox.test(rbf$test_aucs, rbf$cv_aucs)

# Bind them together and summarise mean testAUC by model
test_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
  melt_data() %>% 
  select(-variable) %>% 
  group_by(model, Performance) %>% 
  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
  filter(Performance=="testing")

@@

# Bind them together and summarise mean cvAUC by model
cv_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
  melt_data() %>% 
  select(-variable) %>% 
  group_by(model, Performance) %>% 
  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
  filter(Performance=="cross-validation")

# Get the difference between mean cvAUC and testAUC for each model
difference <- cv_all$mean_AUC - test_all$mean_AUC[match(cv_all$model, test_all$model)]
difference_model <- data.frame(difference, test_all$model)
# Get the order index from small to large of the differences
difference_index <- order(difference_model$difference)

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent, accessible and transparent machine learning methods; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase transparent ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which are ML models that increase in complexity respectively. Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model. The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing/splitting data for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-2)` OTUs and FIT) and a two-class label that defines the colonic health of the patients (SRN or normal). We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection. We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1]. For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1]. For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners. We validated the prediction performances of each hyperparameter setting (Figure S1].

@@

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using AUROC as classification performance metric. Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2]. L1 linear SVM had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2]. However, it also had significanly higher performances than L2 linear SVM, RBF SVM, L2 logistic regression and decision tree which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`), `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. To get a sense of the most important features in linear models, we calculated the mean weights of all the features over the 100 data-splits for each model. [Figure 4]. In the three linear models, FIT was the feature that drove the detection of SRNs, followed by eight OTUs which belong to family *Lachnospiraceae* and *Ruminococcaceae* (four with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get the importance of features in a decision tree, we used gini importance which allows us to use the improvement in the split-criterion is the importance measure attributed to the splitting variable

>>>>

```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
######################################################################
#----------------- Read in important functions -----------------#
######################################################################
source("../code/learning/functions.R")

#####################################################################

######################################################################
#----------------- Load OTU table and MetaData -----------------#
######################################################################
source("../code/learning/load_data.R")

######################################################################
#--------- Load .csv data to get mean test AUC for 7 models--------#
######################################################################

# Read in the cvAUCs, testAUCs for 100 splits.
best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
all <- map_df(best_files, read_files)

@@

rf_xgboost <- wilcoxon_test(all, "Random_Forest", "XGBoost")
rf_logit <- wilcoxon_test(all, "Random_Forest", "L2_Logistic_Regression")
rf_L2svm <- wilcoxon_test(all, "Random_Forest", "L2_Linear_SVM")
xgboost_L2svm <- wilcoxon_test(all, "XGBoost", "L2_Linear_SVM")
logit_L2svm <- wilcoxon_test(all, "L2_Logistic_Regression", "L2_Linear_SVM")

# Bind them together and summarise mean testAUC by model
test_all <- all %>%
  melt_data() %>% 
  select(-variable) %>% 
  group_by(model, Performance) %>% 
  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
  filter(Performance=="testing")

@@

# Bind them together and summarise mean cvAUC by model
cv_all <- all %>%
  melt_data() %>% 
  select(-variable) %>% 
  group_by(model, Performance) %>% 
  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
  filter(Performance=="cross-validation")

# Get the difference between mean cvAUC and testAUC for each model
difference <- cv_all$mean_AUC - test_all$mean_AUC[match(cv_all$model, test_all$model)]
difference_model <- data.frame(difference, test_all$model)
# Get the order index from small to large of the differences
difference_index <- order(abs(difference_model$difference))

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Most notably, there are misleading practices of using "leaky" ML pipelines where models are tested on training data or reporting the best outcome of different iterations of cross-validation. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase reliable ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which are ML models that increase in complexity respectively. Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model. The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))` (std ± `r paste(round(test_all[performance_index[1],4], 3))`) to `r paste(round(test_all[performance_index[7],3], 3))` (std ± `r paste(round(test_all[performance_index[7],4], 3))`). Random forest had the highest mean AUROC for detecting SRN. Despite the simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal). All the cases were independently labeled through colonoscopies. We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection. We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1]. For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1]. For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners. We validated the cross-validation performances of each hyperparameter setting [Figures S1 and S2].

@@

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits in terms of the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric. Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2]. L2 logistic regression and XGBoost had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2]. Random forest also had significanly higher performances than L2 linear SVM, RBF SVM and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data. We calculated the mean weights of all the features over training the each model 100 times [Figure 3a]. In the three linear models, OTUs that drove the detection of SRNs belong to family *Lachnospiraceae* and *Ruminococcaceae* (OTUs..). To explain the importance of features in non-linear models, we used permutation importance on the held-out test data [Figure 3b].
